---
title: "Big Data Analytics"
subtitle: 'Lecture 8:<br>Visualization'
author: "Prof. Dr. Ulrich Matter"
output:
  ioslides_presentation:
    css: ../style/ioslides.css
    template: ../style/nologo_template.html
logo: ../img/logo.png
bibliography: ../references/bigdata.bib
---


```{r set-options, echo=FALSE, cache=FALSE, purl=FALSE}
options(width = 100)
library(knitr)
```


# Updates



# (Big) Data Visualization


## `ggplot2`

- 'Grammar of Graphics'
- Build plots layer-by-layer
- *Here*: Useful tool for exploratory visualization
- In-memory operations
     - Works well with 1 million obs.



## Load taxi trip data


```{r warning=FALSE, echo=TRUE, message=FALSE}

# SET UP----
# see 05_aggregtion_visualization.Rmd for details
# load packages
library(data.table)
library(ggplot2)

# import data into RAM (needs around 200MB)
taxi <- fread("../data/tlc_trips.csv",
              nrows = 100000)

```

## Data preparation

```{r}
# remove unused variables
taxi$V8 <- NULL
taxi$V9 <- NULL
# set covariate names according to the data dictionary
# see https://www1.nyc.gov/assets/tlc/downloads/pdf/data_dictionary_trip_records_yellow.pdf
# note instead of taxizonne ids, long/lat are provided

varnames <- c("vendor_id",
              "pickup_time",
              "dropoff_time",
              "passenger_count",
              "trip_distance",
              "start_long",
              "start_lat",
              "dest_long",
              "dest_lat",
              "payment_type",
              "fare_amount",
              "extra",
              "mta_tax",
              "tip_amount",
              "tolls_amount",
              "total_amount")
names(taxi) <- varnames
# clean the factor levels
taxi$payment_type <- tolower(taxi$payment_type)
taxi$payment_type <- factor(taxi$payment_type, levels = unique(taxi$payment_type))     
```


## Visualization of spatial data with `ggplot2`

- Data source: NYC Taxi & Limousine Commission (TLC).
- Data on all trip records including *pick-up and drop-off times/locations*.


## Preparations

- Load packages for GIS data/operations

```{r message=FALSE, warning=FALSE}
# load GIS packages
library(rgdal)
library(rgeos)
```

## Download map data

```{r message=FALSE, warning=FALSE}
# download the zipped shapefile to a temporary file, unzip
URL <- "https://www1.nyc.gov/assets/planning/download/zip/data-maps/open-data/nycd_19a.zip"
tmp_file <- tempfile()
download.file(URL, tmp_file)
file_path <- unzip(tmp_file, exdir= "../data")
# delete the temporary file
unlink(tmp_file)

```

## Import map data

```{r message=FALSE, warning=FALSE}
# read GIS data
nyc_map <- readOGR(file_path[1], verbose = FALSE)
# have a look at the polygons that constitute the map
summary(nyc_map)
```


## Change map projection

```{r}
# transform the projection
nyc_map <- spTransform(nyc_map, 
                       CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"))
# check result
summary(nyc_map)
```

## Prepare map for plotting with `ggplot2`

```{r warning=FALSE, message=FALSE}
nyc_map <- fortify(nyc_map)
```


## Prepare pick-up and drop-off data


```{r}
# taxi trips plot data
taxi_trips <- taxi[start_long <= max(nyc_map$long) & 
                        start_long >= min(nyc_map$long) &
                        dest_long <= max(nyc_map$long) &
                        dest_long >= min(nyc_map$long) &
                        start_lat <= max(nyc_map$lat) & 
                        start_lat >= min(nyc_map$lat) &
                        dest_lat <= max(nyc_map$lat) &
                        dest_lat >= min(nyc_map$lat) 
                        ]
taxi_trips <- taxi_trips[sample(nrow(taxi_trips), 50000)]

```


## Code time dimension(s)

```{r}
taxi_trips$start_time <- hour(taxi_trips$pickup_time)
```


```{r}
# define new variable for facets
taxi_trips$time_of_day <- "Morning"
taxi_trips[start_time > 12 & start_time < 17]$time_of_day <- "Afternoon"
taxi_trips[start_time %in% c(17:24, 0:5)]$time_of_day <- "Evening/Night"
taxi_trips$time_of_day  <- factor(taxi_trips$time_of_day, levels = c("Morning", "Afternoon", "Evening/Night"))

```


## Base plot: Map of NYC


```{r}
# set up the canvas
locations <- ggplot(taxi_trips, aes(x=long, y=lat))
# add the map geometry
locations <- locations + geom_map(data = nyc_map,
                                  map = nyc_map,
                                  aes(map_id = id))
locations
```

## Add pick-up locations

```{r}
# add pick-up locations to plot
locations + 
     geom_point(aes(x=start_long, y=start_lat),
                color="orange",
                size = 0.1,
                alpha = 0.2)

```

## Add drop-off locations

```{r}
# add pick-up locations to plot
locations +
     geom_point(aes(x=dest_long, y=dest_lat),
                color="steelblue",
                size = 0.1,
                alpha = 0.2) +
     geom_point(aes(x=start_long, y=start_lat),
                color="orange",
                size = 0.1,
                alpha = 0.2)
 

```


## Taxi traffic over the course of a day

```{r fig.height=3, fig.width=9}

# pick-up locations 
locations +
     geom_point(aes(x=start_long, y=start_lat),
                color="orange",
                size = 0.1,
                alpha = 0.2) +
     facet_wrap(vars(time_of_day))
 
```

## Taxi traffic over the course of a day

```{r fig.height=3, fig.width=9}

# drop-off locations 
locations +
     geom_point(aes(x=dest_long, y=dest_lat),
                color="steelblue",
                size = 0.1,
                alpha = 0.2) +
     facet_wrap(vars(time_of_day))
 
```


## Taxi traffic over the course of a day

```{r}
# drop-off locations 
locations +
     geom_point(aes(x=dest_long, y=dest_lat, color = start_time ),
                size = 0.1,
                alpha = 0.2) +
     scale_colour_gradient2( low = "red", mid = "yellow", high = "red",
                             midpoint = 12)
 
```








## References {.smaller}

<style>
slides > slide { overflow: scroll; }
slides > slide:not(.nobackground):after {
  content: '';
}
</style>


